HMM

### 2015-11-06 paper3.Rnw
fixing issues

Given a sequence of animal movements observed via direct observation
or telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
-- formalized in ecological terms as the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} -- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behavior is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin_quantitative_1998}.

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We take data from 
\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous hidden semi-Markov
models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)

@@

\cite{van_de_kerk_hidden_2014} found that the best-fitting HSMMs
according to BIC incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
van_de_kerk_hidden_2014}.As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{van_de_kerk_hidden_2014}'s model-based approach to
identifying the number of behavioural modes.

Our second goal, therefore, is to explore whether
\citeauthor{van_de_kerk_hidden_2014}'s results might be driven at
least in part by structural problems with the HMM, i.e.  the
assumption of temporal homogeneity in behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting relatively
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

Hidden semi-Markov models (HSMM), an extension of HMM that permits
explicit modelling of dwell times (ie. Time spent in a specific state)
using alternative distributions \citep{langrock_flexible_2012} were
used to reveal multiphasic movement of Florida panthers
\citep{van_de_kerk_hidden_2014}. Of 18 Florida panthers GPS-tracked in
hourly intervals, 13 panthers were modelled independently with 2 to 6
states time homogeneous HSMM model with alternative distributions for
step lengths (log-Normal, gamma, and Weibull), turning angle (von
Mises and wrapped Cauchy) and dwell times (Poisson and negative
binomial). Weibull and wrapped Cauchy were the choice of distributions
chosen via Akaike information criterion (AIC) among the alternative
distributions for step length and turning angle
respectively. Furthermore, Bayesian information criterion (BIC) were
used to evaluate goodness of fit between models with different number
of states. BIC values of these models decrease as the number of states
increased from three to six states suggesting the six states model was
favoured statistically, on the other hand, it was hard to interpret
the six states biologically. Despite the flexibility to model dwell
time, it did not affect results, thus, in this paper, we will use HMM
instead of HSMM. Furthermore, the state occupancies based on the
Viterbi algorithm, which predicts the most likely sequence of movement
states based on observed movement results
\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
future movement patterns and cannot capture which movement modes
panthers are likely to occupy depending on time of day from the time
homogeneous transitions (ie, the transition between all states are
constant through time).

@@

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{van_de_kerk_hidden_2014} for HMM fitting. BIC was used for model selection because we were interested in explaining the data generation process by identifying the 'true' number of underlying movement states.

@@

The package depmixS4 \citep{visser2012package} for R
(R Development Core Team 2011) was used for fitting covariate dependent transition HMM
models, simulating states and step lengths using the
estimated parameters, and for estimating states with the Viterbi
algorithm.

@@

Identifying behavioural states based on some set of observations is a common methodological problem in behavioural ecology; finding solutions that are both biologically and statistically interpretable and defensible is challenging. Models that suggest a higher number of states fit the data well but are hard to interpret biologically; models that are biologically interpretable are not statistically supported. While the HMM framework has been applied in many research areas, very few studies have considered modeling with more than 3 states. HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding. For example, \citet{van_de_kerk_hidden_2014} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitat distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.

>>>>

Given a sequence of animal movements observed via direct observation
or telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
-- formalized in ecological terms as the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} -- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behavior is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin1998quantitative}.

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We take data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
according to BIC incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}.As \citeauthor{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of behavioural modes.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e.  the
assumption of temporal homogeneity in behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting relatively
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

Hidden semi-Markov models (HSMM), an extension of HMM that permits
explicit modelling of dwell times (ie. Time spent in a specific state)
using alternative distributions \citep{langrock_flexible_2012} were
used to reveal multiphasic movement of Florida panthers
\citep{kerk2015hidden}. Of 18 Florida panthers GPS-tracked in
hourly intervals, 13 panthers were modelled independently with 2 to 6
states time homogeneous HSMM model with alternative distributions for
step lengths (log-Normal, gamma, and Weibull), turning angle (von
Mises and wrapped Cauchy) and dwell times (Poisson and negative
binomial). Weibull and wrapped Cauchy were the choice of distributions
chosen via Akaike information criterion (AIC) among the alternative
distributions for step length and turning angle
respectively. Furthermore, Bayesian information criterion (BIC) were
used to evaluate goodness of fit between models with different number
of states. BIC values of these models decrease as the number of states
increased from three to six states suggesting the six states model was
favoured statistically, on the other hand, it was hard to interpret
the six states biologically. Despite the flexibility to model dwell
time, it did not affect results, thus, in this paper, we will use HMM
instead of HSMM. Furthermore, the state occupancies based on the
Viterbi algorithm, which predicts the most likely sequence of movement
states based on observed movement results
\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
future movement patterns and cannot capture which movement modes
panthers are likely to occupy depending on time of day from the time
homogeneous transitions (ie, the transition between all states are
constant through time).

@@

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting. BIC was used for model selection because we were interested in explaining the data generation process by identifying the 'true' number of underlying movement states.

@@

The package depmixS4 \citep{visser2010depmixs4} for R
(R Development Core Team 2011) was used for fitting covariate dependent transition HMM
models, simulating states and step lengths using the
estimated parameters, and for estimating states with the Viterbi
algorithm.

@@

Identifying behavioural states based on some set of observations is a common methodological problem in behavioural ecology; finding solutions that are both biologically and statistically interpretable and defensible is challenging. Models that suggest a higher number of states fit the data well but are hard to interpret biologically; models that are biologically interpretable are not statistically supported. While the HMM framework has been applied in many research areas, very few studies have considered modeling with more than 3 states. HMM is a simple straight forward framework, but HMM extensions of adding a small amount of complexity can complicate the model greatly and reveal deeper understanding. For example, \citet{kerk2015hidden} used 2 to 6 movement states HSMM to model Florida panthers,   \citet{morales_extracting_2004} used habitat distance dependent transition HMM to model elk, and outside of ecology, \citet{raffa_multivariate_2015} used mixed effect transition and emission HMM in a smoking study.
